---
title: "T1D_RNAseq"
output: html_document
date: "2026-02-04"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This document is for the analysis of TCR gene usage in RNAseq data from individuals with RNA seq. We will pull raw data from ___ RNA seq datasets and then process them to extract sequencing info, raw counts including TCR genes, HLA genotypes, and MixCR clones (probably very few)

Datasets:
- Dufort et al., https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE124400
    Large dataset containing a compilation of four other studies with similar collection methdologies, the study is based on treated patients but we can probably ignore that. The dataset contains 138 subjects, all with T1D confirmed at 4 different sample points. It also contains information on age, ethnicity/race, c peptide and 
- DAISY, https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE230370
    Large dataset containing the 


```{r,eval=FALSE}
#Code for running from raw sequencing data
#usage
#Process.sh   name_to_file.tsv                    Raw_dir                     Outdir                TmpDir
#-p paired end dataset
# bash ./src/Process.sh ./Data/Dufort/Dufort.tsv /data/Raw/T1D/RNAseq/Dufort/ ./Data/Dufort/ /data/Temp/robben/RNA/dufort/
# bash ./src/Process.sh ./Data/Daisy/Daisy.tsv /data/Raw/T1D/RNAseq/Daisy/ ./Data/Daisy/ /data/Temp/robben/RNA/daisy/ --p
# ###
# bash ./src/Process.sh ./Data/Codazzi/Codazzi.tsv /data/Raw/T1D/RNAseq/Codazzi/ ./Data/Codazzi/ /data/Temp/robben/RNA/codazzi/ 
# bash ./src/Process.sh ./Data/Lorenzon/Lorenzon.tsv /data/Raw/T1D/RNAseq/Lorenzon/ ./Data/Lorenzon/ /data/Temp/robben/RNA/lorenzon/ --p

```


#Set up

```{r}
setwd("~/Projects/RNAseq/") #Change to actual directory
```

```{r}
require(ggplot2)
require(dplyr)
require(plyr)
require(pheatmap)
require(umap)
require(tidyr)
require(data.table)
require(cowplot)
require(Matrix)
require(EDASeq)
require(DESeq2)
library(ggheatmap)
require(ComplexHeatmap)
require(sva)
library(ggrepel)
require(ggpubr)
library(EnhancedVolcano)
require(clusterProfiler)
require(DOSE)
library(DGEobj.utils)
library(rtracklayer)
require(glmnet)
require(lme4)
require(caret)
require(SCGLR)
require(pROC)
require(enrichplot)
require(Seurat)
# install.packages("remotes")
# remotes::install_github("icbi-lab/immunedeconv")


# BiocManager::install("org.Hs.eg.db", character.only = TRUE)
library("org.Hs.eg.db", character.only = TRUE)

pub <- theme(text = element_text(size = 20))
pubrot <-theme(text = element_text(size = 20),axis.text.x = element_text(angle = 45, hjust = 1))
```



#Metadata

```{r read in meta}
#######################
#####Daisy
daisymeta <- read.table(".//Data/Daisy/DaisyMeta.csv",header = T,sep = ",") #Age (age_ia), 
head(daisymeta)

dim(daisymeta) #551 files
length(unique(daisymeta$ID))#143 patients #21 control patients
table(daisymeta$ID) #Looks like 2-7 samples (mean of 4)
table(daisymeta$ID,daisymeta$Group)
rownames(daisymeta) <- daisymeta$Run
daisysex <- rep("male", nrow(daisymeta))
daisysex[daisymeta$female_yn == 1] <- "female"
daisysex[is.na(daisymeta$female_yn)] <- NA
daisydisease <- rep("T1D",nrow(daisymeta))
# daisydisease[daisymeta$Group == "Progressor"] <- "T1D"
daisydisease[daisymeta$Group %in% c("Maintainer","Reverter")] <- "Pre-Diab"
daisydisease[daisymeta$Group %in% c("","Control")] <- "Healthy"

daisyaab <- rep("Pos",nrow(daisymeta))
daisyaab[daisymeta$Group %in% c("Control")] <- "Neg"


hist(unique(daisymeta$age_ia)) #Average age of 7 looks like for initial?
ggplot(daisymeta[daisymeta$Group != "",],aes(y = ID,x = AGE,color = Group)) +
  geom_point() +
  geom_line(aes(group = ID)) 

#Maintainer is AAB+, no T1D
#Reverter loses AAB+, no T1D
#Progresser is AAB+, goes T1D
# daisymeta[daisymeta$Group %in% c('',]$Group <- "Control"
T1D <- rep("control",nrow(daisymeta))
T1D[daisymeta$Group %in% c("Progressor","Maintainer")] <- "case"
daisymeta <- within(daisymeta,Status <- T1D)

#######################
#####Dufort

dufort <- read.table(".//Data/Dufort/DufortMetaData.csv",header = T,sep = ",")
head(dufort)
#age_at_enrollment, Race, rate_of_c.peptide_change, sex, source_name, subject, visit_day (age+visit?), lymphocyte percent (if want to predict t cell amount)
table(dufort$subject) #about 4 per subject
#all t1d?

########################
####codazzi
codazzi <- read.table(".//Data/Codazzi/CodazziMeta.csv",header = T,sep = ",")
head(codazzi)
table(codazzi$genotype) #all diab
# cell_type, Center.Name, tissue, Bases


######################
#####Lorenzon
lorenzon <- read.table(".//Data/Lorenzon/LorenzonMetadata.csv",header = T,sep = ",")
head(lorenzon)
table(lorenzon$patient_diagnosis) #43 healthy, 39 t1d
#age, Bases, sex, tissue, 
lorenzondiag <- lorenzon$patient_diagnosis
lorenzondiag[lorenzondiag == "healthy"] <- "Healthy"
lorenzondiag[lorenzondiag == "type 1 diabetes"] <- "T1D"
lorenstatus <- rep("case",nrow(lorenzon))
lorenstatus[lorenzon$patient_diagnosis == "healthy"] <- "control"
#######################
####Get a common table
#Features: 
#ID (need a unique ID per patient)
#Age
#Geo
#Sex
#Tissue (allblood?)
#Replicate (not super important)
#Disease
#AAB (def not in every)
#Seq depth
metadata <- rbindlist(
list(
  Daisy = data.frame(Run = daisymeta$Run, ID = daisymeta$ID, Age_at_pos = daisymeta$age_ia, 
                     Age_at_collection = daisymeta$AGE, Sex = daisysex,Geo_country = rep("USA",nrow(daisymeta)),
                     Geo_state = rep("CO",nrow(daisymeta)),Tissue = daisymeta$tissue,Clinical_status = daisydisease,
                     AAB = daisyaab, depth = daisymeta$Bases,Status = T1D),
  Dufort = data.frame(Run = dufort$Run, ID = dufort$subject, Age_at_pos = dufort$age_at_enrollment, 
                      Age_at_collection = dufort$age_at_enrollment + (dufort$visit_day/365), Sex = dufort$sex,Geo_country = rep("USA",nrow(dufort)),
                      Geo_state = rep("WA",nrow(dufort)),Tissue = rep("blood",nrow(dufort)),Clinical_status = rep("T1D",),
                      AAB = rep("Pos",nrow(dufort)), depth = dufort$Bases, Status = rep("case",nrow(dufort))),
  Codazzi = data.frame(Run = codazzi$Run, ID = codazzi$BioSample, Age_at_pos = rep(NA,nrow(codazzi)), 
                      Age_at_collection = rep(NA,nrow(codazzi)), Sex = rep(NA,nrow(codazzi)),Geo_country = rep("Italy",nrow(codazzi)),
                      Geo_state = rep("Milano",nrow(codazzi)),Tissue = rep("blood",nrow(codazzi)),Clinical_status = rep("T1D",nrow(codazzi)),
                      AAB = rep("Pos",nrow(codazzi)), depth = codazzi$Bases,Status = rep("case",nrow(codazzi))),
  Lorenzon = data.frame(Run = lorenzon$Run, ID = lorenzon$Sample.Name, Age_at_pos = rep(NA,nrow(lorenzon)), 
                      Age_at_collection = lorenzon$age, Sex = lorenzon$sex,Geo_country = rep("France",nrow(lorenzon)),
                      Geo_state = rep("Paris",nrow(lorenzon)),Tissue = rep("blood",nrow(lorenzon)),Clinical_status = lorenzondiag,
                      AAB = rep(NA,nrow(lorenzon)), depth = lorenzon$Bases, Status = lorenstatus)
  
),idcol = "Study")

head(metadata)
table(metadata[!duplicated(metadata$ID)]$Study,metadata[!duplicated(metadata$ID)]$Clinical_status)
metadata <- as.data.frame(metadata)
rownames(metadata) <- metadata$Run
#Plot out the studies
ggplot(metadata[!is.na(metadata$Age_at_collection),],aes(x = Age_at_collection,y = ID,colour = Clinical_status)) +
  geom_point() +
  geom_line() +
  ylab("Patient ID") +
  theme(axis.text.y = element_blank()) +
  pub
ggsave(filename = ".//Results/Fig/Fig4StudyAge.pdf",device = "pdf",width = 5,height = 20,limitsize = F)

stats <- rep("control",nrow(metadata))
stats[metadata$Clinical_status == "T1D" | (metadata$Clinical_status == "Pre-Diab" & metadata$AAB == "Pos")] <- "case"
metadata <- within(metadata,Status <- stats)
metadata <- as.data.frame(metadata)
rownames(metadata) <- metadata$Run

meta <- metadata[!duplicated(metadata$ID),]
table(meta$Sex)
table(meta$Geo_country)
table(meta$Clinical_status,meta$Status)
table(metadata$AAB,metadata$Status)

#Finally make the metadata plot to show how many patients in each study are of contition and status
ggplot(metadata,aes(x = Study,fill = Status)) +
  geom_bar() +
  ylab("# of samples") +
  xlab("Study") +
  pubrot

#Plot age
ggplot(metadata,aes(x = Age_at_collection,fill = Study))+
  geom_density(alpha = 0.5) +
  facet_wrap(~Status,ncol = 1) +
  geom_vline(xintercept = 12,linetype = "dashed") +
  xlab("Age") +
  pub

write.table(metadata,file = ".//Data/metadata.tsv")
```


#Expression Data

Read in the expression data, which is in a different file per sample.

```{r Read in expression data}
#We need to write a function that will get the data into a basic matrix (probably don't need an expr set)
dirs <- as.list(c(".//Data/Daisy/Counts/",".//Data/Dufort/Counts/",".//Data/Codazzi/Counts/",".//Data/Lorenzon/Counts/"))
files <- unlist(lapply(dirs,function(x) paste(x,list.files(x),sep="")))

readExpr <- function(files){
  # files <- paste(dir,list.files(dir),sep = "")
  i <- 1
  for (f in files){
    name <- unlist(lapply(strsplit(basename(f),"_"),function(x) x[[1]]))
    df <- read.table(f,row.names = 1,header = F,sep = "\t")
    colnames(df) <- name
    if(i == 1){
      expr <- df
    } else {
      expr <- cbind(expr,df)
    }
    i <- i+1
  }
  print(paste("Read ",i," files",sep=""))
  return(expr)
}
exp <- readExpr(files)
head(exp)
dim(exp)
exp <- exp[rowSums(exp) > 1,]
dim(exp) #Filtered 34773 genes
depth <- colSums(exp)
plot(depth)
metadata <- cbind(metadata,data.frame(Sum_Count = depth[metadata$Run]))
exp <- exp[,metadata$Run]
# ggplot(metadata,aes(x = Run,y = Sum_Count,color = Study)) +
#   geom_point()



#Data exploration
samped <- unlist(lapply(split(metadata,metadata$Study),function(x) rownames(x)[1:10]))
plotRLE(as.matrix(exp)[,samped])
norm <-  vst(as.matrix(exp))
plotRLE(as.matrix(norm)[,samped])

#PCA
pca <- prcomp(t(norm))
tmp <- cbind(metadata,pca$x[,1:2])
ggplot(tmp[,!colnames(tmp) == "Sum_Count"],aes(x = PC1,y = PC2, colour = Status, shape = Study)) +
  geom_point(size = 2,alpha = 0.5) +
  scale_x_continuous(limits = c(-50,100)) +
  scale_y_continuous(limits = c(-50,50)) +
  pub

#TCR genes
Tgenes <- rownames(exp)[grep(c("^TR[A|B|G|D][V|D|J|C][0-9]"),rownames(norm))]
TCRgenes <- norm[Tgenes,]
TCRgenes <- TCRgenes[rowSums(TCRgenes) > 0,] #removed about 50 not expressed genes
TCRgenes <- TCRgenes[,colSums(TCRgenes) > 0] #removed about 50 not expressed genes
dim(TCRgenes) #243 genes all 1430 patients

boxplot(TCRgenes)
plotRLE(as.matrix(TCRgenes))

#Plot TCR genes
sorder <- rev(order(apply(TCRgenes,1,mean)))
genecol <- factor(substr(rownames(TCRgenes)[sorder],1,4))
library(RColorBrewer)
num_levels <- nlevels(genecol)
factor_colors <- brewer.pal(num_levels, "Set3")

boxplot(t(TCRgenes)[,sorder],col = factor_colors[genecol])
legend("topright", legend = levels(genecol), col = factor_colors, pch = 15)

# #Save the expression data
write.table(TCRgenes,file = ".//Data/TCRexp.tsv",sep = "\t",row.names = T)
```


#HLA Data

```{r read in HLA function}
readHLA <- function(x) {
  files <- list.files(x)
  f <- files[grep("genotype",files)]
  f <- paste(x,f,sep="/")
  df <- read.csv(f,header = F,sep = "\t")
  colnames(df) <- c("gene_name","num_diff_alleles","allele_1","abundance_1","quality_1",
                    "allele_2","abundance_2","quality_2","secondary_alleles")
  return(df)
}
dirs <- as.list(c(list.dirs(".//Data/Daisy/HLA/")[-1],list.dirs(".//Data/Dufort/HLA/")[-1],list.dirs(".//Data/Codazzi/HLA/")[-1],list.dirs(".//Data/Lorenzon/HLA/")[-1]))
hla.ls <- lapply(dirs,function(x) readHLA(x))
#Now write a function that will get them all out
uniqueHLA <- unlist(lapply(hla.ls,function(x) strsplit(c(x$allele_1,x$allele_2),",")))
uniqueHLA <- unique(uniqueHLA[uniqueHLA != "."])
getHLAtable <- function(x,hlas){
  mat <- matrix(0,ncol=length(hlas))
  colnames(mat) <- hlas
  rep <- matrix(0,ncol = nrow(x))
  colnames(rep) <- x[,1]
  for(i in 1:nrow(x)){
    row <- x[i,]
    detalleles <- row[2]
    allele1 <- row[3:5]
    a1name <- gsub(",*","",allele1[1])
    allele2 <- row[6:8]
    a2name <- gsub(",*","",allele2[1])
    if (detalleles == 0) {
      next
    } else if(detalleles == 1) {
      rep[,i] <- 1
      if(allele1[3] > 5 & a1name %in% hlas) {
        mat[,a1name] <- 2
      }
    } else {
      rep[,i] <- 2
      if(allele1[3] > 5 & a1name %in% hlas ) {
        mat[,a1name] <- 1
      }
      if(allele2[3] > 5 & a2name %in% hlas) {
        mat[,a2name] <- 1
      }
    }
  }
  return(list(mat,rep))
}
hla.mat <- lapply(hla.ls,function(x) getHLAtable(x,uniqueHLA))

hla <- do.call(rbind,lapply(hla.mat,function(x) x[[1]]))
rep <- do.call(rbind,lapply(hla.mat,function(x) x[[2]]))
names <- basename(unlist(dirs))
rownames(hla) <- names
rownames(rep) <- names
dim(hla)
hla <- hla[,colSums(hla) > 0]
hla <- hla[rowSums(hla) > 0,]
dim(hla) # Lost about half of hlas and 3 patients who had no predicted for some reason
rep <- rep[rowSums(rep) > 0,]
write.table(x = hla,file = ".//Data/HLA.tsv",sep = "\t")
#Allele completion by sequencing depth

complete <- rep/rep
complete[is.nan(complete)] <- 0
complete <- rowSums(complete)

plot(complete,depth[names(complete)],ylab = "Sequencing depth (reads)",xlab = "Number of HLA detected")
mean(complete)
sd(complete)
#Let's make sure there is consistency of samples by ID since allele should not change
calcError <- function(x,df){
  # print(dim(df[x,]))
  # print(x)
  test <- colSums(df[x,])
  return(1 - sum(tmp[x,test < length(x)])/sum(df))
}

patientfiles <- split(metadata$Run,metadata$ID)
patientfiles <- patientfiles[unlist(lapply(patientfiles,length)) > 1]
ptdepth <- unlist(lapply(patientfiles,function(x) sum(depth[x])))
lapply(patientfiles,function(x) length(x))
tmp <- hla/hla
tmp[is.nan(tmp)] <- 0
error <- do.call(cbind,lapply(patientfiles,function(x) tryCatch(calcError(x,tmp),error = function(e) { return(tmp[1]/0)})))
enames <- colnames(error)[!error == Inf]
error <- error[!error == Inf]
names(error) <- enames
plot(error,ptdepth[enames],xlab = "Error",ylab = "Sequencing depth (reads)")
mean(error)
sd(error)

#Look at the effect of HLA on different categories
hla_merge <- cbind(hla,metadata[rownames(hla),]) #466 which matches the IDs

#Get a per patient HLA or else we are artificially increasing OR
pthla <- hla[!duplicated(metadata[rownames(hla),]$ID),]

sum.df <- data.frame(NULL)
for (x in colnames(pthla)){
  tmp <- pthla[,colnames(pthla) == x]
  tmp <- metadata[rownames(pthla)[tmp > 0],]
  sum.df <- rbind(sum.df,data.frame(Total = nrow(tmp), Case = nrow(tmp[tmp$Status == "case",]),Control = nrow(tmp[tmp$Status == "control",])))
}
rownames(sum.df) <- colnames(pthla)



calcOR <- function(x,row){
  #         allele  rest
  # health  a      b
  #   t1d      c      d
  allele <- as.numeric(x[row,])
  notallele <- as.numeric(colSums(x[-row,]))
  # print(allele)
  # print(notallele)
  
  data_table <- matrix(c(allele[2], notallele[2], allele[3], notallele[3]), ncol = 2, byrow = TRUE)
  # print(data_table)
  # Calculate the odds ratio using fisher.test()
  fisher_result <- fisher.test(data_table)
  odds_ratio_value <- fisher_result$estimate
  clneg <- fisher_result$conf.int[1]
  clpos <- fisher_result$conf.int[2]
  return(data.frame(OR = odds_ratio_value,Min = clneg, Max = clpos))
}

ORsum <- cbind(sum.df,rbindlist(lapply(as.list(1:nrow(sum.df)),function(x) calcOR(sum.df,x))))
ORsum[rev(order(summary$OR)),]
```


Figure out if calcualted OR is equal to predicted OR

```{r Known vs Calc OR}

#Finally use the defined protective and risk alleles
riskalleles <- read.table(file = ".//HLA/Risk_Alleles.txt",header = T,sep = "\t")
riskalleles <- riskalleles[-58,]
riskalleles <- riskalleles[!riskalleles$Allele.Haplotype == "",]
riskalleles$Allele <- paste("HLA-",riskalleles$Allele,sep="")

extractOR <- function(x){
  # print(x)
  string <- iconv(as.character(x[4]), from = "ISO-8859-1", to = "UTF-8")
  split <- unlist(strsplit(string,"\\\n"))
  split <- gsub("OR ","",split)
  OR <- as.numeric(unlist(lapply(strsplit(split," "),function(x) x[[1]])))
  range <- unlist(lapply(strsplit(split," "),function(x) x[[2]]))
  min <- NULL
  max <- NULL
  for (r in range){
    if (r == "specified") {
      min <- c(min,NA)
      max <- c(max,NA)
    }else if (substr(r,1,2) == "(P"){
      min <- c(min,NA)
      max <- c(max,NA)
    } else {
      min <- c(min,as.numeric(gsub("\\(","",unlist(lapply(strsplit(r,"-",fixed = T),function(x) x[[1]])))))
      max <- c(max,as.numeric(gsub("\\)*","",unlist(lapply(strsplit(r,"-",fixed = T),function(x) x[[2]])))))
    }
  }
  return(cbind(as.data.frame(x[c(1,3,5,6,7)][rep(1,length(OR)),]),data.frame(OR = OR, min = min, max = max)))
}
allelelist <- rbindlist(lapply(split(riskalleles,riskalleles$Allele),function(x) extractOR(x)))


#Compare to calculated

ORsum <- within(ORsum, Base <- lapply(strsplit(rownames(ORsum),":"),function(x) paste(x[1:min(length(x),2)],collapse=":")))
tsum <- ORsum[!unlist(apply(ORsum,1,function(x) any(is.infinite(x$OR)))),]
tsum <- tsum[!tsum$OR == 0,]
tsum <- rbind(within(tsum,Set <- rep("Calculated",nrow(tsum))),
              data.frame(Total = rep(NA,nrow(allelelist)),Case = rep(NA,nrow(allelelist)),Control = rep(NA,nrow(allelelist)), 
               OR = allelelist$OR,Min = allelelist$min, Max = allelelist$max,Base = allelelist$Allele,Set = rep("Known",nrow(allelelist)))
)
tsum <- tsum[!is.na(tsum$OR),]
tmp <- split(tsum,as.character(tsum$Base))
tmp <- lapply(tmp,function(x) rbindlist(lapply(split(x,x$Set),function(y) y[rev(order(y$OR)),])))
tsum <- rbindlist(tmp[rev(order(unlist(lapply(tmp,function(x) mean(x$OR)))))])
tsum <- within(tsum,Name <- rev(c(1:nrow(tsum))))
tdf <- tsum
tdf$OR <- (sign(log2(tsum$OR))* (2^(abs(log2(tsum$OR)))))#log2(tdf$OR) * 2
tdf$Min <- (sign(log2(tsum$Min))* (2^(abs(log2(tsum$Min)))))
tdf$Max <- (sign(log2(tsum$Max))* (2^(abs(log2(tsum$Max)))))
tdf$Base <- as.character(tdf$Base)

hack <- data.frame(Gene = unique(tdf$Base), Sig = unlist(lapply(split(tdf,tdf$Base),function(x) mean(x$OR) > 2 | mean(x$OR) < -2)), cols = rep("black",length(unique(tdf$Base))))

hack[!hack$Sig,3] <- "blue"

p <- ggplot(tdf,aes(y = Name,color = Set)) +
  geom_vline(xintercept = 0,linetype = "dashed",color = "grey",alpha = 1) +
  geom_point(aes(x=OR),size = 2) +
  # geom_linerange(aes(xmin=sapply(Min,function(x) max(-40,x)), xmax=sapply(Max,function(x) min(40,x)))) +
  geom_linerange(aes(xmin=Min, xmax=Max),linewidth = 1) +
  facet_wrap(~factor(Base,levels = unique(Base)),ncol = 1,strip.position = "left",space = "free_y",scales = "free_y") +
  scale_y_discrete(expand = c(0, 1)) +
  
  scale_x_continuous(limits = c(-15,15)) +
  xlab("Log2 normalized OR") +
  theme(strip.clip = "off",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),panel.spacing.y = unit(0.7, "lines"),
        strip.text.y.left = element_text(angle = 0),
        axis.line = element_line(colour = "black"),
        strip.text = element_text(color = hack$cols)
        )
ggsave(filename = ".//Results/Fig/HLA_OR_small.pdf",device = "pdf",width = 5,height = 100,limitsize = F)

#Also plot the best and worst
ggplot(tdf[tdf$Base %in% unique(tdf$Base)[c(1:10,276:286)],],aes(y = Name,color = Set)) +
  geom_vline(xintercept = 0,linetype = "dashed",color = "grey",alpha = 1) +
  geom_point(aes(x=OR),size = 2) +
  geom_linerange(aes(xmin=Min, xmax=Max),linewidth = 1) +
  facet_wrap(~factor(Base,levels = unique(Base)),ncol = 1,strip.position = "left",space = "free_y",scales = "free_y") +
  scale_y_discrete(expand = c(0, 1)) +
  
  scale_x_continuous(limits = c(-25,25)) +
  xlab("Log2 normalized OR") +
  theme(strip.clip = "off",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),panel.spacing.y = unit(0.7, "lines"),
        strip.text.y.left = element_text(angle = 0),
        axis.line = element_line(colour = "black")
        )

write.table(tdf,file = ".//Results/Table/Supptable_HLA_risk.tsv",sep = "\t",quote = F)
```

Categorize the alleles

```{r Categorize}

agg_OR <- aggregate(OR ~ Base,data = tdf, FUN = mean)
#Classify each individual by HLA risk
or <- 6
risk <- agg_OR[agg_OR$OR > or,]$Base#unique(allelelist[allelelist$OR > 5,]$Allele)
prot <- agg_OR[agg_OR$OR < -or,]$Base#unique(allelelist[allelelist$OR < 0.1,]$Allele)

#Using 2 is 95 prot, 10 risk most both
#Using 3 is 221 prot, 435 risk 694 both
#Using 4 is 332 prot, 321 risk, 454 both
#Using 5 is 300 prot, 364 risk, 366 both #This gives the greatest spread

has_risk <- rowSums(hla[,unlist(lapply(risk,function(x) which(x == substr(colnames(hla),1,nchar(x)))))]) >= 1
has_prot <- rowSums(hla[,unlist(lapply(prot,function(x) which(x == substr(colnames(hla),1,nchar(x)))))]) >= 1
risk.df <- data.frame(has_risk,has_prot)
risky <- rep("Neither",nrow(risk.df))
risky[has_risk] <- "Risk"
risky[has_prot] <- "Protective"
risky[has_prot & has_risk] <- "Both"
table(risky)
risk.df <- within(risk.df, Risk <- risky)

metadata <- within(metadata,Risk <- risk.df[rownames(metadata),]$Risk)
table(metadata$Risk,metadata$Status)
submeta <- metadata[!duplicated(metadata$ID),]
table(submeta$Risk,submeta$Status)

#Plot the OR > 4
ggplot(tdf[tdf$Base %in% c(prot,risk),],aes(y = Name,color = Set)) +
  geom_vline(xintercept = 0,linetype = "dashed",color = "grey",alpha = 1) +
  geom_point(aes(x=OR),size = 2) +
  geom_linerange(aes(xmin=Min, xmax=Max),linewidth = 1) +
  facet_wrap(~factor(Base,levels = unique(Base)),ncol = 1,strip.position = "left",space = "free_y",scales = "free_y") +
  scale_y_discrete(expand = c(0, 1)) +
  scale_x_continuous(limits = c(-25,25)) +
  xlab("Log2 normalized OR") +
  theme(strip.clip = "off",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.spacing.y = unit(0.1, "lines"),
        strip.text.y.left = element_text(angle = 0),
        axis.line = element_line(colour = "black"),
        text = element_text(size = 10)
        )

orplot <- data.frame(OR = 1:5,
           Both = c(1424,1317,694,454,366), 
           Risk = c(2,10,435,320,364), 
           Prot = c(1,95,221,332,300), 
           Neither = c(0,5,77,321,397)) %>% pivot_longer(cols = c("Both","Risk","Prot","Neither"))
ggplot(orplot,aes(x = OR,y = value,color = name,group = name)) +
  geom_line(size = 2) +
  xlab("OR greater than") +
  ylab("# of samples") +
  pub

write.table(metadata,file = ".//Data/metadata.tsv")
```



#MixCR

Lets load in the CDR3 data from each patient, likely it will be low and heavily influenced by sequencing depth

```{r read in CDR3}
#Read in the cdr3 from each
dirs <- c(list.dirs(".//Data/Daisy/MixCR/")[-1],list.dirs(".//Data/Dufort/MixCR/")[-1],
          list.dirs(".//Data/Codazzi/MixCR/")[-1],list.dirs(".//Data/Lorenzon/MixCR/")[-1])
dirlist <- as.list(dirs)
names(dirlist) <- basename(dirs)
# getCDR3 <- function(x) {
#   # dirs <- list.dirs(x)[-1]
#   # dirlist <- as.list(dirs)
#   # names(dirlist) <- basename(dirs)
#   flist <- lapply(dirlist,readCDR3)
#   return(rbindlist(flist,idcol = "ID",fill = T))
# }

readCDR3 <- function(y){
  ls <- as.list(paste(y,"/clones_",c("IGH","IGK","IGL","TRA","TRB","TRD","TRG"),".tsv",sep = ""))
  names(ls) <- c("IGH","IGK","IGL","TRA","TRB","TRD","TRG")
  ls <- lapply(ls,function(x) if (file.exists(x)){ read.csv(x,header = T,sep = "\t") })
  ls <- lapply(ls,function(x) x[,colnames(x) %in% c("cloneId","readCount","aaSeqCDR3",
                                                    "allVAlignments","allDAlignments","allJAlignments","allCAlignments")])
  df <- rbindlist(ls,idcol = "Type")
  return(df)
}

# studies <- list(Daisy = ".//Data/Daisy/MixCR/")

mix <- rbindlist(lapply(dirlist,readCDR3),idcol = "ID",fill = T)

# cdr3 <- mix[[1]]
```

Analyze the CDR3

```{r analyze CDR3}
dim(mix) #2.2 million entries
table(mix$Type)
tcr <- cdr3[cdr3$Type %in% c("TRA","TRB","TRD","TRG")]
tcr_table <- as.data.frame(table(tcr$ID,tcr$Type))
ggplot(tcr_table,aes(x = Var2, y = Freq,fill = Var2)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.5) +
  xlab("TCR gene class") +
  ylab("# of detected CDR3") +
  guides(fill = F) +
  pub
#Average of 200-300 clones per sample

trab <- cdr3[cdr3$Type %in% c("TRA","TRB")]
seqs <- unique(trab$aaSeqCDR3)
samplenames <- unique(trab$ID)
seqmat <- sparseMatrix(i = as.numeric(as.factor(trab$aaSeqCDR3)),j = as.numeric(as.factor(trab$ID)),x = trab$readCount)
dimnames(seqmat) <- list(levels(as.factor(trab$aaSeqCDR3)),levels(as.factor(trab$ID)))
total <- colSums(seqmat)
ggplot(data.frame(TCR = total,depth = depth[colnames(seqmat)]),aes(y = total,x = depth)) +
  geom_point() +
  geom_smooth(method = "lm",se = F)+
  xlab("Sequencing Depth (reads)") +
  ylab("Total RNA-seq CDR3 sequences") +
  pubrot
ggsave(filename = ".//Results/Fig/mixcr_sd.svg",width = 4,height = 4)

###############################################
#######

#So obviously this is not enough CDR3 sequences to even do anything, let's check the VDJdb annotations

```


#Differential expression

Lets first test the differential expression between disease affected individuals to see if there is a difference

```{r}
dds <- DESeqDataSetFromMatrix(countData = exp,
                              colData = metadata,
                              design= ~ Status)
dds
dds <- DESeq(dds)
dds
resultsNames(dds)
res <- results(dds, name="Status_control_vs_case")
write.table(res,".//Results/Table/SupptableDEGres.tsv",sep = "\t")
plotMA(res)
EnhancedVolcano(res,
    lab = rownames(res),
    x = 'log2FoldChange',
    y = 'pvalue')
sigres <- res[res$padj < 0.01 & abs(res$log2FoldChange) > 2,]
# or to shrink log fold changes association with condition:
# res <- lfcShrink(dds, coef="condition_trt_vs_untrt", type="apeglm")

#Try to set random intercept for Study and sex
dds <- DESeqDataSetFromMatrix(countData = exp,
                              colData = metadata,
                              design= ~Status+(1|Study))
dds
dds <- DESeq(dds)
dds
resultsNames(dds)
res <- results(dds, name="Status_control_vs_case")
write.table(res,".//Results/Table/SupptableDEGadjres.tsv",sep = "\t")
plotMA(res)
EnhancedVolcano(res,
    lab = rownames(res),
    x = 'log2FoldChange',
    y = 'pvalue')
sigres <- res[res$padj < 0.01 & abs(res$log2FoldChange) > 2,]

```

Take a look at the gsea

```{r}

original_gene_list <- sigres$log2FoldChange
names(original_gene_list) <- rownames(sigres)
gene_list<-na.omit(original_gene_list)
gene_list = sort(gene_list, decreasing = TRUE)
gse <- gseGO(geneList=gene_list, 
             ont ="ALL", 
             keyType = "SYMBOL", 
             minGSSize = 3, 
             maxGSSize = 800, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = "org.Hs.eg.db", 
             pAdjustMethod = "none")
gse@result

dotplot(gse, split=".sign") + facet_grid(.~.sign)
dotplot(gse)#, split=".sign") + facet_grid(.~.sign)
```

#Deconvolution

```{r}
library(GenomicFeatures)
# gff_data <- import("/data/Databases/Genomes/GRCh38_p14/genomic.gtf")
txdb <- txdbmaker::makeTxDbFromGFF("/data/Databases/Genomes/GRCh38_p14/genomic.gtf", format = "gtf")
exons <- exonsBy(txdb, by = "tx")
gene_lengths <- sum(width(reduce(exons)))

TPM <- convertCounts(countsMatrix = counts_df, 
                      unit = "tpm", 
                      geneLength = gene_lengths, 
                      log = FALSE, 
                      normalize = "none")
decon <- immunedeconv::deconvolute(exp, "quantiseq",tumor = F)

```



#TCR Normalization

We want to take a look at the general differences in the TCR gene expression

```{r Normalize TCR genes}
TCRraw <- exp[Tgenes,]
dim(TCRraw) #243 genes all 1430 patients

plot(colSums(matrix(TCRraw > 0,nrow(TCRraw),ncol(TCRraw))))

plotRLE(as.matrix(TCRraw)[,samped],main = "Raw")
plotRLE(as.matrix(TCRgenes)[,samped],main = "vst")
#Are C genes normal?

means <- colMeans(TCRgenes[grep("C",rownames(TCRgenes)),])
sds <- colSds(TCRgenes[grep("C",rownames(TCRgenes)),])
plot(means, pch=19, xlab="Run", ylab="Expression", xaxt="n", 
     main="TCR vst constant expression", col="red", cex=.3)
segments(x0 = 1:length(means),y0 = means - sds, x1 = 1:length(means),y1 = means + sds, 
       col="black",lwd = 0.1)
#Looks like there is some variation in C gene expression, can we normalize against it like a housekeeping gene?
#Maybe just do SCT transformation which uses a negative binomial to estimate factors including C genes


TCRsct <- sctransform::vst(as.matrix(TCRraw),cell_attr = metadata,batch_var = "Study")$y
plotRLE(as.matrix(TCRsct)[,samped],main = "SCT")


means <- colMeans(TCRsct[grep("C",rownames(TCRsct)),])
sds <- colSds(as.matrix(TCRsct[grep("C",rownames(TCRsct)),]))
plot(means, pch=19, xlab="Run", ylab="Expression", xaxt="n", 
     main="TCRsct constant expression", col="red", cex=.3)
segments(x0 = 1:length(means),y0 = means - sds, x1 = 1:length(means),y1 = means + sds, 
       col="black",lwd = 0.1)

sctorder <- rev(order(apply(TCRsct,1,mean)))
genecol <- factor(substr(rownames(TCRsct)[sctorder],1,4))
library(RColorBrewer)
num_levels <- nlevels(genecol)
factor_colors <- brewer.pal(num_levels, "Set3")
boxplot(t(TCRsct)[,sctorder],col = factor_colors[genecol],main = "SCT TCR normalization")
legend("topright", legend = levels(genecol), col = factor_colors, pch = 15)
############Looks like the best will be to use the VST (variance is normalized across batch) and then norm against mean expression of C genes)
means <- colMeans(TCRgenes[grep("C",rownames(TCRgenes)),])
sds <- colSds(TCRgenes[grep("C",rownames(TCRgenes)),])
TCRnorm <- sweep(TCRgenes, 2, means, "-")

means <- colMeans(TCRnorm[grep("C",rownames(TCRnorm)),])
sds <- colSds(as.matrix(TCRnorm[grep("C",rownames(TCRnorm)),]))
plot(means, pch=19, xlab="Run", ylab="Expression", xaxt="n", 
     main="TCRnorm constant expression", col="red", cex=.3,ylim = c(-0.5, 0.5))
segments(x0 = 1:length(means),y0 = means - sds, x1 = 1:length(means),y1 = means + sds, 
       col="black",lwd = 0.1)

plotRLE(as.matrix(TCRnorm)[,samped],main = "c gene norm")
boxplot(t(TCRnorm)[,sorder],col = factor_colors[genecol])
legend("topright", legend = levels(genecol), col = factor_colors, pch = 15)

#ComBat normalization
mod <- model.matrix(~as.factor(Status), data=metadata)
TCRclean <- ComBat(dat=TCRraw, batch=as.factor(metadata$Study), mod=mod)
means <- colMeans(TCRclean[grep("C",rownames(TCRgenes)),])
sds <- colSds(TCRclean[grep("C",rownames(TCRgenes)),])
plot(means, pch=19, xlab="Run", ylab="Expression", xaxt="n", 
     main="TCR ComBat constant expression", col="red", cex=.3)
segments(x0 = 1:length(means),y0 = means - sds, x1 = 1:length(means),y1 = means + sds, 
       col="black",lwd = 0.1)
plotRLE(as.matrix(TCRclean)[,samped],main = "ComBat norm")
boxplot((t(TCRclean)[,sorder]),col = factor_colors[genecol])
legend("topright", legend = levels(genecol), col = factor_colors, pch = 15)

#Compare SCT to sequencing depth
plot(colMeans(TCRraw),metadata$depth,col = factor(metadata$Study),xlab = "Mean TCR gene",ylab = "Sequencing depth")
plot(colMeans(TCRsct),metadata$depth,col = factor(metadata$Study),xlab = "Mean TCR SCT",ylab = "Sequencing depth")

write.table(x = TCRsct,file = ".//Data/TCRnorm.tsv",sep = "\t",quote = F)
```

#TCR Exploration

```{r exploring TCR genes}

###heatmap
# ggheatmap(TCRgenes,cluster_cols = T,cluster_rows = T,
#           annotation_cols = metadata[,c(1,6,10,11,13,15)],annotation_color = )
TCRscaled <- t(apply(TCRsct, 1, scale))
Heatmap(TCRscaled,top_annotation = HeatmapAnnotation(Status = metadata$Clinical_status,Study = metadata$Study, Sex = metadata$Sex,
                                                     Age = anno_barplot(metadata$Age_at_collection)))


#Do basic Dimensionality reduction
umap <- umap(t(TCRsct))
umap <- umap$layout
colnames(umap) <- c("umap1","umap2")
metadata <- within(within(metadata,umap1 <- umap[rownames(metadata),1]),umap2 <- umap[rownames(metadata),2])

ggplot(metadata,aes(x = umap1,y = umap2,color = Study,group = ID)) +
  geom_point() +
  ggtitle("TCR SCT")
ggplot(metadata,aes(x = umap1,y = umap2,color = Clinical_status,group = ID,shape = AAB)) +
  geom_point()
ggplot(metadata,aes(x = umap1,y = umap2,color = Risk,group = ID)) +
  geom_point()



#Let's initially just look at differences in each based on status
combined <- cbind(metadata, t(TCRsct)[rownames(metadata),])
long <- combined %>% pivot_longer(grep("TR",colnames(combined),value = T),names_to = "TCRgene",values_to = "Exp")

ggplot(long,aes(x = TCRgene,y = Exp,fill = Status)) +
  stat_summary(fun = "mean",geom = "bar",position = "dodge") + 
  pubrot


```


#TCR Differential

```{r}
#We can't look at differential expression modeled by a negative binomial because the data has been transformed to a negative binomial and variance stabilized
#or can we?
#Make a function that will t.test based on the TCR gene per subject based on status
performTest <- function(dat, col, adjust = 1){
  x <- dat[dat$Status == "case",]$Exp
  y <- dat[dat$Status == "control",]$Exp
  if ((length(x) < 2) | length(y) < 2){
    return(NA)
  } 
  logfc <- log2(mean(x)/mean(y))
  pval <- t.test(x,y)$p.value
  padj <- pval * adjust
  return(data.frame(LogFC = logfc, case = mean(x), control = mean(y), pvalue = pval, p.adj = padj))
}

sp <- split(long, long$TCRgene)
dif <- rbindlist(lapply(sp,function(x) performTest(x,1,length(sp))),idcol = "TCRgene")
sig <- dif[abs(dif$LogFC) > 1 & dif$p.adj < 0.05,]
write.table(dif,".//Results/Table/Suppfig9DiffDET.tsv",sep="\t")

#Volcano plot
cols <- rep("black",nrow(dif))
cols[abs(dif$LogFC) > 1] <- "blue"
cols[dif$p.adj < 0.05] <- "orange"
cols[abs(dif$LogFC) > 1 & dif$p.adj < 0.05] <- "red"

p1 <- ggplot(within(dif,col <- cols),aes(x = LogFC,y = -log10(p.adj))) +
  geom_point(color = cols) +
  geom_hline(yintercept = 1.3) +
  geom_vline(xintercept = c(-1,1)) +
  geom_text_repel(data = dif[cols == "red",], # Subset the data for the text layer
            aes(label = dif[cols == "red",]$TCRgene)) +
  pub

#####Terminal diff expression
Termdif <- rbindlist(lapply(sp,function(x) performTest(x[x$Age_at_collection > 12,],1,length(sp))),idcol = "TCRgene")

#Now create a plot with just the significant and status
# p2 <- ggplot(long[long$TCRgene == sig$TCRgene,],aes(x = TCRgene,y = Exp,fill = Status)) +
#   geom_boxplot() +
#   pubrot

p2 <- ggplot(long[long$TCRgene %in% sig$TCRgene & !is.na(long$Age_at_collection),],aes(x = TCRgene,y = Exp,fill = Status)) +
  geom_boxplot() +
  facet_wrap(~Age_at_collection > 11,labeller = as_labeller(c(
    `FALSE` = "Age < 12",
    `TRUE`   = "Age > 12"
  )))+
  scale_y_continuous(limits = c(0,10))+
  pubrot
p2
ggarrange(p1,p2,widths = c(0.5,1))
ggsave(filename = ".//Results/Fig/Fig4TCRdet.svg",width = 10,height = 5,limitsize = F)

############################Also split by risk?
sp_risk<- split(long, paste(long$TCRgene,long$Risk,sep=" "))
dif_risk <- lapply(sp_risk,function(x) performTest(x,1,length(sp_risk)))
dif_risk <- rbindlist(dif_risk[!is.na(dif_risk)],idcol = "Gene_Risk")
dif_risk <- within(dif_risk, TCRgene <- unlist(lapply(strsplit(dif_risk$Gene_Risk," "),function(x) x[[1]])))
dif_risk <- within(dif_risk, Risk_group <- unlist(lapply(strsplit(dif_risk$Gene_Risk," "),function(x) x[[2]])))

sig_risk <- dif_risk[abs(dif_risk$LogFC) > 1 & dif_risk$p.adj < 0.01,]
print(sig_risk)
write.table(dif_risk,".//Results/Table/Suppfig10RiskDETDiff.tsv",sep="\t")

cols <- rep("black",nrow(dif_risk))
cols[abs(dif_risk$LogFC) > 1] <- "blue"
cols[dif_risk$p.adj < 0.05] <- "orange"
cols[abs(dif_risk$LogFC) > 1 & dif_risk$p.adj < 0.05] <- "red"
p1 <- ggplot(within(dif_risk,col <- cols),aes(x = LogFC,y = -log10(p.adj))) +
  geom_hline(yintercept = 1.3,colour = "grey") +
  geom_vline(xintercept = c(-1,1),colour = "grey") +
  geom_point(color = cols) +
  geom_text_repel(data = dif_risk[cols == "red",], # Subset the data for the text layer
            aes(label = dif_risk[cols == "red",]$TCRgene)) +
  facet_wrap(~Risk_group,ncol = 1,scales = "free") +
  pub
p1
#Now create a plot with just the significant and status
# p2 <- ggplot(long[long$TCRgene == sig$TCRgene,],aes(x = TCRgene,y = Exp,fill = Status)) +
#   geom_boxplot() +
#   pubrot

p2 <- ggplot(long[long$TCRgene %in% sig_risk$TCRgene,],aes(x = TCRgene,y = Exp,fill = Status)) +
  geom_boxplot() +
  facet_wrap(~Risk,ncol = 1)+
  pubrot
p2
ggarrange(p1,p2,widths = c(0.5,1))


#Plot some of the interesting over time
ggplot(long[long$TCRgene %in% sig$TCRgene,],aes(x = Age_at_collection,y = Exp,colour = Status,group = Status)) +
  geom_smooth(method = "loess") +
  facet_wrap(~TCRgene)

ggplot(long[long$TCRgene == "TRBV7-6",],aes(x = Age_at_collection,y = Exp,colour = Status,group = Status)) +
  geom_smooth(method = "loess") +
  ylab("Expression") +
  xlab("Age") +
  ggtitle("TRBV7-6 Expression") +
  pub

```




#TCR Model

Lets simply ask is there any TCR gene that change over time in any groups specific

```{r Allele affect}
allele <-  "Protective"
names <- rownames(risk.df[risk.df$Risk == allele,])
# names <- colnames(TCRsct)
umap <- umap(t(TCRsct[,names]))
umap_data <- data.frame(
  UMAP1 = umap$layout[, 1],
  UMAP2 = umap$layout[, 2],
  metadata[names,]
)
a <- ggplot(umap_data[umap_data$Age_at_collection < 12,], aes(x = UMAP1, y = UMAP2, color = Status)) +
  geom_point(size = 5,alpha = 0.8) + # Adjust point size as needed
  labs(title = paste("Expression in",allele,"Alleles",sep=" "),subtitle = "under 12") +
  theme_minimal() + pub
b <- ggplot(umap_data[umap_data$Age_at_collection > 12,], aes(x = UMAP1, y = UMAP2, color = Status)) +
  geom_point(size = 5,alpha = 0.8) + # Adjust point size as needed
  labs(title = paste("Expression in",allele,"Alleles",sep=" "),subtitle = "over 12" ) +
  theme_minimal() + pub
ggarrange(a,b,ncol = 1)
# umap_data <- within(umap_data, strange <- umap_data$UMAP2 < -5)


```

Not really any affect by allele unfortunate lets model the genes we know

```{r}
ggplot(long[long$TCRgene %in% sig$TCRgene,],aes(x = Age_at_collection, y = Exp,color = Status,group = Status ))+
  # geom_point() +
  geom_smooth() +
  facet_wrap(~TCRgene,nrow = 1) +
  ggtitle("Significant TCR genes by status") +
  ylab("TCR expression") +
  xlab("Age") +
  scale_x_continuous(labels = c(10,50),breaks = c(10,50)) +
  pub

#Compare to 5 random TCR genes
ggplot(long[long$TCRgene %in% Tgenes[runif(5,0,200)],],aes(x = Age_at_collection, y = Exp,color = Status,group = Status ))+
  # geom_point() +
  geom_smooth() +
  facet_wrap(~TCRgene,nrow = 1) +
  ggtitle("Random TCR genes by status") +
  pub


#Take a look at the risk group individuals
ggplot(long[long$TCRgene %in% sig_risk$TCRgene & long$Risk %in% c("Both","Protective","Risk"),],aes(x = Age_at_collection, y = Exp,color = Status,group = Status ))+
  # geom_point() +
  geom_smooth() +
  scale_x_continuous(limits = c(0,20),breaks = c(5,20)) +
  facet_grid(TCRgene~Risk,scales = "free") +
  ggtitle("Allele Significant TCR genes by status") +
  xlab("Age") +
  ylab("TCR relative expression") +
  theme(strip.text.y.right = element_text(angle = 0),
        axis.text.y = element_blank()) +
  pub
```




```{r}
# #Now plot by age and TCR gene
# merge <- cbind(umap_data,t(TCRsct))
# merge_long <- merge %>%
#   pivot_longer(rownames(TCRsct),names_to = "TCR_gene",values_to = "Expression")
# 
# names <- unique(merge_long$TCR_gene)
# j <- 10
# ggplot(merge_long[merge_long$TCR_gene == names[j],],aes(x = Age,y = Expression,color = strange)) +
#   geom_smooth() +
#   ggtitle(names[j]) 
#   
# 
# #So we want to maybe make a stacked barplot of TCR genes with the average or number in the y and interpolated over age make a function that will pull from a set and #interpolate over age
# 
# 
# interpTCR <- function(x,df,ages){
#   sub <- df[df$TCR_gene == x,]
#   if(nrow(sub) < 1){
#     return(NA)
#   }
#   loess_fit <- loess(Expression ~ Age, data = sub[!is.na(sub$Age),], span = 0.5)
#   preds <- predict(loess_fit,newdata = ages,se = T)
#   return(data.frame(Age = ages,Interp = preds$fit, SE = preds$se.fit))
# }
# 
# genelist <- as.list(rownames(TCRnorm))
# names(genelist) <- rownames(TCRnorm)
# inter <- rbindlist(lapply(genelist,function(x) interpTCR(x,merge_long,seq(from = 3, to = 20, by = 1))),idcol = "TCRgene")
# 
# ggplot(inter, aes(x=Age, y=Interp, fill=TCRgene)) + 
#     geom_area() +
#   guides(fill = FALSE)
# 
# 
# #Separate by Group
# group_inter <- rbindlist(
# list(
#   # control = rbindlist(lapply(genelist,function(x) interpTCR(x,merge_long[merge_long$Group == "Control",],seq(from = 3, to = 20, by = 1))),idcol = "TCRgene"),
#   maintainer = rbindlist(lapply(genelist,function(x) interpTCR(x,merge_long[merge_long$Group == "Maintainer",],seq(from = 3, to = 20, by = 1))),idcol = "TCRgene"),
#   reverter = rbindlist(lapply(genelist,function(x) interpTCR(x,merge_long[merge_long$Group == "Reverter",],seq(from = 3, to = 20, by = 1))),idcol = "TCRgene"),
#   progressor = rbindlist(lapply(genelist,function(x) interpTCR(x,merge_long[merge_long$Group == "Progressor",],seq(from = 3, to = 20, by = 1))),idcol = "TCRgene")
# ),idcol = "Group"
# )
# 
# ggplot(group_inter, aes(x=Age, y=Interp, fill=TCRgene)) + 
#   geom_area() +
#   facet_wrap(~Group,ncol = 3) +
#   guides(fill = F) +
#   ylab("Interpolated Count") +
#   pub 
# ggsave(".//Results/Fig/age_tcrcount.pdf",width = 5,height = 3)
# 
# 
# ggplot(group_inter,aes(x = Age, y = Interp,color = Group,group = Group)) +
#   stat_summary(fun = "mean",geom = "line") +
#   stat_summary(fun = "mean",geom = "point") +
#   stat_summary(fun.data = "mean_se",geom = "errorbar")+
#   ggtitle("Average normalized TCR clonality") +
#   pub
# 
# #by strange
# group_inter <- rbindlist(
# list(
#   # control = rbindlist(lapply(genelist,function(x) interpTCR(x,merge_long[merge_long$Group == "Control",],seq(from = 3, to = 20, by = 1))),idcol = "TCRgene"),
#   strange = rbindlist(lapply(genelist,function(x) interpTCR(x,merge_long[merge_long$strange == T,],seq(from = 3, to = 20, by = 1))),idcol = "TCRgene"),
#   normal = rbindlist(lapply(genelist,function(x) interpTCR(x,merge_long[merge_long$strange == F,],seq(from = 3, to = 20, by = 1))),idcol = "TCRgene")
# ),idcol = "Strange"
# )
# 
# ggplot(group_inter, aes(x=Age, y=Interp, fill=TCRgene)) + 
#   geom_area() +
#   facet_wrap(~Strange,ncol = 1) +
#   guides(fill = F)
# 
# #By status
# group_inter <- rbindlist(
# list(
#   # control = rbindlist(lapply(genelist,function(x) interpTCR(x,merge_long[merge_long$Group == "Control",],seq(from = 3, to = 20, by = 1))),idcol = "TCRgene"),
#   maintainer = rbindlist(lapply(genelist,function(x) interpTCR(x,merge_long[merge_long$Group == "Maintainer",],seq(from = 3, to = 20, by = 1))),idcol = "TCRgene"),
# ),idcol = "Status"
# )
# 
# #By risk
# 
# group_inter <- rbindlist(
# list(
#   risk = rbindlist(lapply(genelist,function(x) interpTCR(x,merge_long[merge_long$Risk == "Risk",],seq(from = 3, to = 20, by = 1))),idcol = "TCRgene"),
#   protective = rbindlist(lapply(genelist,function(x) interpTCR(x,merge_long[merge_long$Risk == "Protective",],seq(from = 3, to = 20, by = 1))),idcol = "TCRgene"),
#   both = rbindlist(lapply(genelist,function(x) interpTCR(x,merge_long[merge_long$Risk == "Both",],seq(from = 3, to = 20, by = 1))),idcol = "TCRgene")
# ),idcol = "Risk"
# )
# 
# ggplot(group_inter, aes(x=Age, y=Interp, fill=TCRgene)) + 
#   geom_area() +
#   facet_wrap(~Risk,ncol = 1) +
#   guides(fill = FALSE)
# 
# ggplot(group_inter,aes(x = Age, y = Interp,color = Risk,group = Risk)) +
#   stat_summary(fun = "mean",geom = "line") +
#   stat_summary(fun = "mean",geom = "point") +
#   stat_summary(fun.data = "mean_se",geom = "errorbar")+
#   ggtitle("Average normalized TCR clonality") +
#   pub
# 
# #Check which TCR are different between 
# model <- aov(lm(Expression ~ Risk + Age,data = merge_long))
# summary(model)
# TukeyHSD(model)
# #perform wilcox at each time point between protective and risk
# ager <- as.numeric(cut(merge_long$Age, breaks = 20,1:20))
# results <- data.frame(NULL)
# i <- 1
# for (y in unique(merge_long$TCR_gene)){
#   print(i)
#   for (x in 1:20){
#     sub <- merge_long[ager == x & merge_long$TCR_gene == y,]
#     sub <- sub[!is.na(sub$ID),]
#     risk <- sub[sub$Risk == "Risk",]
#     prot <- sub[sub$Risk == "Protective",]
#     if (nrow(risk) == 0 | nrow(prot) == 0) {
#       next
#     }
#     riskmean <- mean(risk$Expression)
#     protmean <- mean(prot$Expression)
#     p <- wilcox.test(risk$Expression,prot$Expression)$p.value
#     lfc <- log2( riskmean / protmean ) #risk over prot
#     results <- rbind(results,data.frame(Age = x, TCRgene = y, rmean = riskmean, pmean = protmean,LogFC = lfc,p = p, padj = p*length(TCRgenes))) 
#   }
#   i <- i+1
# }
# sig <- unique(results[abs(results$LogFC) > 1 & results$p < 0.05, ]$TCRgene)
# sig <- sig[!is.na(sig)]
# #specific
# plot.ls <- list()
# for (s in sig){
#   res <- results[results$TCRgene == s,]
#   res <- res[res$p < 0.05,]
#   sub <- group_inter[group_inter$TCRgene == s & group_inter$Risk != "both",]
#   maxes <- max(sub[sub$Age %in% res$Age,]$Interp,na.rm = T)
#   plot.ls[[s]] <-  ggplot(sub,aes(x = Age, y = Interp,color = factor(Risk,levels = c("risk","protective")),group = factor(Risk,levels = c("risk","protective")))) +
#     # stat_summary(fun = "mean",geom = "line") +
#     # stat_summary(fun = "mean",geom = "point") +
#     # stat_summary(,geom = "errorbar") +
#     geom_line() +
#     geom_point() +
#     geom_errorbar(aes(ymin = Interp - SE,ymax = Interp + SE)) +
#     annotate("text",label = "*",x = res$Age,y = maxes + 0.5 ,size = 7) +
#     pub +
#     theme(legend.title = element_blank())+
#     ggtitle(s)
# }
# 
# #Previous identified important
# plot.ls['TRBV23-1']
# plot.ls['TRBV12-1']
```


```{r old list}
# prot <- c(
#   "HLA-A*11:01",
#   "HLA-A*66:01",
#   "HLA-A*32:01",
#   "HLA-B*35:02",
#   "HLA-B*46:01",
#   "HLA-B*57:01",
#   "HLA-B*07:02",
#   "HLA-B*44:03",
#   "HLA-B*57:03",
#   "HLA-DRB1*03:02",
#   "HLA-DRB1*04:03",
#   "HLA-DRB1*08:04",
#   "HLA-DRB1*11:01",
#   "HLA-DRB1*11:04",
#   "HLA-DRB1*13:03",
#   "HLA-DRB1*14:01",
#   "HLA-DRB1*15:01",
#   "HLA-DRB1*15:03",
#   "HLA-DRB1*16:02",
#   "HLA-DQA1*01:03",
#   "HLA-DQA1*02:01",
#   "HLA-DQA1*04:01",
#   "HLA-DQA1*05:05",
#   "HLA-DQB1*03:01",
#   "HLA-DQB1*04:02",
#   "HLA-DQB1*05:01",
#   "HLA-DQB1*05:03",
#   "HLA-DQB1*06:01",
#   "HLA-DQB1*06:02",
#   "HLA-DPB1*04:01",
#   "HLA-DPB1*04:02"
# )
#  
# risk <- c(
#   "HLA-A*02:01",
#   "HLA-A*24:02",
#   "HLA-A*33:03",
#   "HLA-B*08:01",
#   "HLA-B*15:10",
#   "HLA-B*18:01",
#   "HLA-B*39:06",
#   "HLA-B*54:01",
#   "HLA-C*03:03",
#   "HLA-C*03:04",
#   "HLA-C*15:02",
#   "HLA-DRB1*03:01",
#   "HLA-DRB1*04:01",
#   "HLA-DRB1*04:02",
#   "HLA-DRB1*04:04",
#   "HLA-DRB1*04:05",
#   "HLA-DRB1*09:01",
#   "HLA-DPB1*02:02",
#   "HLA-DPB1*03:01"
# )

```









#Linear Modelling (as in TCR profiling)

```{r}
##############Functions
getCoefs <- function(l){
  best_coefs <- coef(l)#, s = lim$lambda.min) #min lambda
  coef_df <- data.frame(Term = rownames(best_coefs),Coef = as.numeric(best_coefs))
  coef_df <- coef_df[rev(order(coef_df$Coef)),]
  coef_df$Term <- factor(coef_df$Term,levels = coef_df$Term)
  return(coef_df)
}
get_ROC <- function(x,y){
  proc_obj <- roc(predictor = x,response = y,ci=TRUE, ci.alpha=0.9, stratified=FALSE,
                  plot=F, auc.polygon=TRUE, max.auc.polygon=TRUE, grid=TRUE,
                  print.auc=TRUE, show.thres=TRUE,quiet = TRUE)
  net_roc <- data.frame(y = rev(proc_obj$sensitivities),x = rev(proc_obj$specificities))
  return(net_roc)
}
get_AUC <- function(y,x){
  if (length(levels(factor(y))) < 2){
    return(NA)
  }
  proc_obj <- roc(predictor = x,response = y,ci=TRUE, ci.alpha=0.9, stratified=FALSE,
                  plot=F, auc.polygon=TRUE, max.auc.polygon=TRUE, grid=TRUE,
                  print.auc=TRUE, show.thres=TRUE,quiet= TRUE)
  net_auc <- auc(proc_obj)
  return(net_auc)
}
tjurs <- function(y,pred){ #Uses the Tjurs psuedo R2 to calculate an error
  probs_failure <- pred[y == 0]
  probs_failure[is.na(probs_failure)] <- 1
  probs_success <- pred[y == 1]
  probs_success[is.na(probs_success)] <- 0
  mean_failure <- mean(probs_failure,na.rm = T)
  mean_success <- mean(probs_success,na.rm = T)
  if (is.nan(mean_failure)){ mean_failure <- 0 }
  if (is.nan(mean_success)){ mean_success <- 0 }
  return(mean_success - mean_failure)
}
zap <- function(x, y){
  stopifnot(length(x) == length(y))
  z <- list()
  for (i in seq_along(x)){
    z[[i]] <- list(x[[i]], y[[i]])
  }
  return(z)
}
getPredictors <- function(y,pred,hla){
  # print(data.frame(y = y, pred = pred))
  acc <- sum(y == (pred > 0.5),na.rm = T)/length(y)
  auc_all <- get_AUC(pred,y)
  r2 <- tjurs(y,pred)
  accuracies <- NULL
  r2s <- NULL
  rocs <- NULL
  aucs <- NULL
  n <- NULL
  for (i in 1:ncol(hla)) {
    rows <- which(hla[,i] == 1)
    # print(rows)
    # print(data.frame(y = y[rows], pred = pred[rows]))
    n <- c(n,length(rows))
    trues <- y[rows] == (pred[rows] > 0.5)
    trues[is.na(trues)] <- FALSE
    #Calc accuracy for hla only
    accs <- sum(trues)/length(rows)
    if(is.nan(accs)) { accs <- 0 }
    accuracies <- c(accuracies,accs)
    #Calc ROC and AUC for hla only
    auc <- get_AUC(y[rows],pred[rows])
    aucs <- c(aucs,auc)
    #Calc R2 for hla only
    r2s <- c(r2s,tjurs(y[rows],pred[rows]))
  }
  return(data.frame(Allele = c('All',colnames(hla)),n = c(length(y),n),Acc = c(acc,accuracies),R2 = c(r2,r2s),AUC = c(auc_all,aucs)))
}
prepFeatures <- function(samp,tcr,alleles,ids){
  ids <- ids[ids %in% colnames(tcr)]
  sub <- samp[ids,]
  response <- ifelse(factor(sub$Status) == "case", 1, 0)
  df <- data.frame(response = response,
                        Age = sub$age_max,
                        alleles[sub$subject_id,],
                        t(scaleTCR)[ids,])
  rownames(df) <- ids
  df <- df[!apply(df,1,function(x) any(is.na(x))),]
  return(df)
}
log.regress <- function(train,test,hla){
  #Train on train
  hla <- hla[rownames(test),]
  lim <- cv.glmnet(as.matrix(train[,2:ncol(train)]),train[,1],family = "binomial",alpha = 1)
  pred <- predict(lim,as.matrix(test[,2:ncol(test)]),s = "lambda.1se",type = "response")
  lim_res <- list(Coefs = getCoefs(lim),Results = getPredictors(test[,1],pred,hla))
  return(lim_res)
}
mv.regress <- function(train,test,hla){
  train_hla <- hla[rownames(train),]
  hla <- hla[rownames(test),]
  k <- t(t(train_hla) * train[,1])
  hla_index <- grep("HLA",colnames(train))
  mv <- multivariateGlm.fit(k,train[,-c(1,hla_index)],family = rep("binomial",ncol(k)),size = NULL)
  coefs <- sapply(mv,coef)
  colnames(coefs) <- colnames(k)
  mvpred <- predict(mv,test[,-c(1,hla_index)],type = "response")
  mvlist <- lapply(mvpred,function(x) getPredictors(test[,1],x,hla))
  allacc <- do.call(cbind,lapply(mvlist,function(x) x$Acc))
  allr2 <- do.call(cbind,lapply(mvlist,function(x) x$R2))
  allauc <- do.call(cbind,lapply(mvlist,function(x) x$AUC))
  mv_res <- list(Coefs = coefs,
                 Inter = data.frame(Term = colnames(coefs),Intercept = coefs[1,]),
                 #as.data.frame(t(data.frame(apply(coefs,1,function(x) t.test(x)$conf.int),row.names = c("CImin","CImax"))))),
                 Results = cbind(mvlist[[1]][,1:2],
                                 data.frame(Acc = apply(allacc,1,mean),
                                            R2 = apply(allr2,1,mean), 
                                            AUC = apply(allauc,1,function(x) mean(x,na.rm=T)))))
  return(mv_res)
  # mv_res <- list(Coefs = coefs,Results = data.frame(Allele = c('All',colnames(hla)),n = c(length(y),n),Acc = c(acc,accuracies),R2 = c(r2,r2s),ROC = c(roc_all,rocs),AUC = c(auc_all,aucs))))
}
make_long <- function(df){
  df <- within(df,sample <- rownames(df))  %>%
    pivot_longer(cols = colnames(df)[grep("HLA",colnames(df))],names_to = "HLA",values_to = "Contains")
  df <- df[df$Contains >= 1,]
  df <- df[,!colnames(df) == "Contains"]
  tmp <- as.data.frame(df)
  colnames(tmp) <- gsub("\\.","_",colnames(tmp))
  tmp[,1] <- as.factor(tmp[,1])
  tcrindex <- grep("TR",colnames(df))
  tmp[,tcrindex] <- apply(tmp[,tcrindex],1,function(x) as.numeric(x)*100000)
  return(tmp)
}
lmm.regress <- function(train,test,hla){
  train.df <- make_long(train)
  test.df <- make_long(test)
  matchhla <- intersect(unique(train.df$HLA),unique(test.df$HLA))
  train.df <- train.df[train.df$HLA %in% matchhla,]
  test.df <- test.df[test.df$HLA %in% matchhla,]
  
  form <- as.formula(paste("response ~", paste(colnames(train.df)[4:ncol(train.df)-2], collapse = " + ")," + (1 | HLA) "))#+ (1 | Age)" )) #Make this on the test set?
  lmm <- glmer(form,data = train.df,
               family = binomial,control = glmerControl(optimizer = "nloptwrap",calc.derivs = FALSE,optCtrl=list(maxfun=2e5)),nAGQ = 0)
  coefs <- coef(lmm)
  coef_table <- summary(lmm)$coefficients
  #Predict
  lmm_pred <- predict(lmm,test.df,type = "response")
  lmm_pred <- split(lmm_pred,test.df$sample)#Get the median prediction 
  lmm_pred <- unlist(lapply(lmm_pred,function(x) mean(as.numeric(x))))
  lmm_res <- list(Coef = data.frame(Term = rownames(coef_table),Coef = coef_table[,1], Error = coef_table[,2],pvalue = coef_table[,4]),
                  Inter = data.frame(Term = rownames(coefs$HLA),Intercept = coefs$HLA[,1]),
                  Results = getPredictors(test[names(lmm_pred),]$response,lmm_pred,hla[names(lmm_pred),]))
  return(lmm_res)
}
###############Normalize and scale dataset
#######################Prep the TCR features
scaleTCR <- apply(TCRnorm,2,scale)
rownames(scaleTCR) <- rownames(TCRnorm)
boxplot(scaleTCR) #Check that scaleTCR is correctly normalized (remove outliers)
plot(apply(scaleTCR,2,median),col = c("red","blue")[d.df[colnames(scaleTCR),]$response + 1])
abline(h = c(-0.24,-0.36), col = "red", lty = 2)
outliers <- which(apply(scaleTCR,2,median) > -0.24 | apply(scaleTCR,2,median) < -0.37)
norm_scaleTCR <- scaleTCR[,-outliers]
################Train test split
d.df <- prepFeatures(subsample,norm_scaleTCR,alleles,rownames(subsample))
hla <- d.df[,grep("HLA",colnames(d.df))]
term.df <- prepFeatures(subsample,norm_scaleTCR,alleles,term_ids)
nterm.df <- prepFeatures(subsample,norm_scaleTCR,alleles,non_term_ids)
equal.df <- prepFeatures(subsample,norm_scaleTCR,alleles,equal_ids)
###############Test lists
runStatTests <- function(train,test,ls,hla){
  hla_index <- ls$hla
  tcr_index <- ls$tcr
  return(list(
    GLM = log.regress(train,test,hla),
    GLM_HLA = log.regress(train[,-tcr_index],test[,-tcr_index],hla),
    GLM_TCR = log.regress(train[,-hla_index],test[,-hla_index],hla),
    MV = mv.regress(train,test,hla),
    LMM = lmm.regress(train,test,hla)
  ))
}

Full_list <- runStatTests(Full$train,Full$test,Full,hla) 
Full_list_noage <- runStatTests(Fullnoage$train,Fullnoage$test,Fullnoage,hla)
Term_list <- runStatTests(Term$train,Term$test,Term,hla) 
nTerm_list <- runStatTests(nTerm$train,nTerm$test,nTerm,hla) 

Term_nTerm_list <- runStatTests(Term$train,nTerm$train,Term,hla) 
nTerm_Term_list <- runStatTests(nTerm$train,Term$train,nTerm,hla) 

Equal_list <- runStatTests(Equal$train,Equal$test,Equal,hla)
Equal_noage <- runStatTests(Equalnoage$train,Equalnoage$test,Equal,hla)
```

#Machine learning

```{r}

##############################################
#### Machine Learning ########################
##############################################

#The problem is that some patients have 0 recorded HLA so we remove those patient files first
features <- data.frame(response = metadata[userows,]$Status, 
                   Age = metadata[userows,]$Age_at_collection,
                   hla[userows,],
                   t(scaleTCR)[userows,])
# features <- d.df
#Now use machine learning code to develop multiple models predicting the response based on TCR and allele usage
#Make a train test set
set.seed(12345)
train_index <- createDataPartition(1:nrow(features), p=0.70, list=FALSE)
train <- features[train_index,]
validation <- features[-train_index,]

trainModels <- function(train,validation){
  train_age <- train$Age
  val_age <- validation$Age
  train <- train[,!colnames(train) == "Age"]
  validation <- validation[,!colnames(validation) == "Age"]
  
  #Iterate through models
  
  control <- trainControl(method="repeatedcv", number=10, repeats=3,classProbs = T)
  
  m_bag <- train(as.factor(response) ~ ., data=train, method="cforest", metric="ROC", 
                 trControl=control, preProcess = c("center", "scale") )
  m_sda <- train(as.factor(response)~., data=train, method="sda", metric="ROC", 
                 trControl=control, preProcess = c("center", "scale") )
  m_mlp <- train(as.factor(response)~., data=train, method="mlp", metric="ROC", 
                 trControl=control, preProcess = c("center", "scale") )
  m_glm <- train(as.factor(response)~., data=train, method="glm", metric="ROC", 
                 trControl=control, preProcess = c("center", "scale") )
  m_boost <- train(as.factor(response)~., data=train, method="gbm", metric="ROC", 
                   trControl=control, preProcess = c("center", "scale") )
  m_svm <- train(as.factor(response)~., data=train, method="svmLinear2", metric="ROC", 
                 trControl=control, preProcess = c("center", "scale") )
  
  # resample_results <- resamples(list(Bag=m_bag,SDA=m_sda, 
                                     # MLP=m_mlp, GLM=m_glm, Boost=m_boost, SVM=m_svm))
  model_list <- list(Bag=m_bag,SDA=m_sda, 
                 MLP=m_mlp, GLM=m_glm, Boost=m_boost, SVM=m_svm)
  # summary(resample_results,metric = c("Kappa","Accuracy"))
  # bwplot(resample_results , metric = c("Accuracy","ROC"))
  
  #Then predict on the validation set 
  y <- rep(0,nrow(validation))#as.character(validation$encoding)
  y[validation$response == "case"] <- 1
  
  p1 <- predict(m_bag, validation, preProcess = c("center", "scale"),type = "prob")
  p2 <- predict(m_sda, validation, preProcess = c("center", "scale"),type = "prob")
  p3 <- predict(m_mlp, validation, preProcess = c("center", "scale"),type = "prob")
  p4 <- predict(m_glm, validation, preProcess = c("center", "scale"),type = "prob")
  p5 <- predict(m_boost, validation, preProcess = c("center", "scale"),type = "prob")
  p6 <- predict(m_svm, validation, preProcess = c("center", "scale"),type = "prob")
  models <- c("Bagging","SDA","MLP","GLM","Boost","SVM")
  accuracies <- c(calcAcc(p1$case,y),calcAcc(p2$case,y),calcAcc(p3$case,y),calcAcc(p4$case,y),calcAcc(p5$case,y),calcAcc(p6$case,y))
  pos_accuracies <- unlist(lapply(list(p1,p2,p3,p4,p5,p6),function(x) calcAcc(x = x$case[y == 1],y = y[y==1])))
  neg_accuracies <- unlist(lapply(list(p1,p2,p3,p4,p5,p6),function(x) calcAcc(x = x$case[y == 0],y = y[y==0])))
  
  plist <- list(bag = p1,sda = p2,mlp = p3,glm = p4,boost = p5,svm = p6)
  plist <- lapply(plist,function(x) within(cbind(validation[,1],val_age,x),acc <- (x$case > 0.05) == (validation$response == "case")))
  
  #ROC
  
  
  predlist <- list(p1,p2,p3,p4,p5,p6)
  names(predlist) <- models
  roclist <- lapply(predlist,function(x) list(roc = get_ROC(x$case,y), auc = get_AUC(x$case,y)))
  names(roclist) <- models
  roc.df <- rbindlist(lapply(roclist,function(x) x[[1]]),idcol = "Model")
  
  a <- ggplot(roc.df,aes(x = x,y = y, colour = Model)) +
    geom_line(linewidth = 1.2) +
    geom_abline(slope = 1, intercept = 1,linetype = "dashed") +
    scale_x_reverse() +
    # scale_x_continuous(expand = c(0,0)) +
    scale_y_continuous(expand = c(0,0)) +
    xlab("Specificity") +
    ylab("Sensitivity") +
    pub
  
  df <- data.frame(Model = models,Accuracy = accuracies,AUC = unlist(lapply(roclist,function(x) x[[2]])),Pos_acc = pos_accuracies, Neg_acc = neg_accuracies)
  return(list(models = model_list,AUC = a,data = df,plist = plist))
}

#Now train for different feature levels
#May need to hide age
hlaonly <- trainModels(train[,-grep("TR",colnames(features))],validation[,-grep("TR",colnames(features))])
tcronly <- trainModels(train[,-grep("HLA",colnames(features))],validation[,-grep("HLA",colnames(features))])
both <- trainModels(train,validation)

#Plot
resample_results <- resamples(hlaonly[[1]])
hla_plot <- bwplot(resample_results , metric = c("Accuracy","ROC"),main = "HLA only")

resample_results <- resamples(tcronly[[1]])
tcr_plot <- bwplot(resample_results , metric = c("Accuracy","ROC"),main = "TCR only")


resample_results <- resamples(both[[1]])
both_plot <- bwplot(resample_results , metric = c("Accuracy","ROC"),main = "HLA + TCR")

par(mfrow = c(1, 3))
hla_plot
tcr_plot
both_plot
par(mfrow = c(1, 1))

#Feature importance 

hlafeatures <- varImp(hlaonly$models$GLM,scale = F)
tcrfeatures <- varImp(tcronly$models$GLM,scale = F)
bothfeatures <- varImp(both$models$GLM,scale = F)

plot(hlafeatures,top=20,main = "HLA features")
plot(tcrfeatures,top=20,main = "TCR features")
plot(bothfeatures,top=20,main = "HLA + TCR features")
```



#Summarization

```{r}
getSummary <- function(x,y) {
  df <- data.frame(Study = x$Study[1],n = length(x$ID), 
                   Female = sum(x$Sex == "female"),
                   Male = sum(x$Sex == "male"),
                  as.data.frame(t(matrix(table(x$Clinical_status)))),
                  as.data.frame(t(matrix(table(x$AAB)))),
                  as.data.frame(t(matrix(table(x$Status)))),
                  age_mean = mean(x$Age_at_collection,na.rm = T), 
                  age_sd = sd(x$Age_at_collection,na.rm = T),
                  depth = mean(y$depth), depthsd = sd(y$depth))
  # colnames(df) <- c("Study","n","female","male","Healthy","PD","D","Neg","Pos","Case","Control",
                    # "age_mean","age_sd","depth","depth_sd")
  return(df)
}

getSummary(meta,metadata)
lapply(split(metadata,metadata$Study),function(x) getSummary(x[!duplicated(x$ID),],x))

getSummaryHT <- function(x,hla,tcr) {
  hlacols <- colnames(hla)[colnames(hla) %in% rownames(x)]
  tcrcols <- colnames(tcr)[colnames(tcr) %in% rownames(x)]
  return(data.frame(meanHLA = mean(colSums(hla[,hlacols],na.rm = T),na.rm = T),
                    sdHLA = sd(colSums(hla[,hlacols],na.rm = T),na.rm = T),
                    meanTCR = mean(colSums(tcr[,tcrcols],na.rm = T),na.rm = T),
                    sdTCR = sd(colSums(tcr[,tcrcols],na.rm = T),na.rm = T)))
}

getSummaryHT(metadata,t(hla),TCRraw > 1)
lapply(split(metadata,metadata$Study),function(x) getSummaryHT(x,t(hla),TCRraw > 1))
```







